---
title: "CLT theorem via exponential distribution"
author: "Kirill Panarin"
output: html_document
---
## Overview

Let's investigate the exponential distribution and compare the application of the Central Limit Theorem on it. The exponential distribution has a rate parameter. I will use rate parameter equal to 0.2 and will assign it to the `lambda` variable. The mean of exponential distribution is `1/lambda` and the standard deviation is also `1/lambda`. In addition to `lambda` we need to define a variable for the theoretical mean and standard deviation:

```{r}
lambda <- 0.2
theoretical.mean <- 1/lambda
theoretical.sd <- 1/lambda
```

## Simulations

The simulation process has the following structure:
- Generate `n` random values.
- Repeat the first procedure for `nsim` number of times in order to gather a reasonable amount of data for analysis.

I will use the following values for `n` and `nsim` parameters:

```{r}
n <- 40
nsim <- 1000
```

Let's start simulations from generating `n` times `nsim` random values with the exponential distribution. `simulation.data_raw` contains the array of random values and `simulation.data_matrix` contains the same data but restructured as a matrix with `n` columns:

```{r}
simulation.data_raw <- rexp(n * nsim, lambda)
simulation.data_matrix <- matrix(simulation.data_raw, ncol = n)
```

The first visualization will be the samples values distribution. It just shows that we are really dealing with the exponential distribution:

```{r, echo = FALSE}
library(ggplot2)
ggplot(data = data.frame(x = simulation.data_raw), aes(x = x, y = ..density..)) + 
    geom_histogram(binwidth = 0.7, color = "black", fill = "orange") +
    geom_density(lwd = 1) +
    ggtitle("Distribution of random exponentials")
```


```{r, echo = FALSE, fig.width=9, fig.height = 6, fig.align='center'}
nosim <- 1000
cfunc <- function(x, n) sqrt(n) * (mean(x) - theoretical.mean) / theoretical.sd
dat <- data.frame(
  x = c(apply(matrix(sample(0:1, prob = c(.1,.9), nosim * 10, replace = TRUE), 
                     nosim), 1, cfunc, 10),
        apply(matrix(sample(0:1, prob = c(.1,.9), nosim * 20, replace = TRUE), 
                     nosim), 1, cfunc, 20),
        apply(matrix(sample(0:1, prob = c(.1,.9), nosim * n, replace = TRUE), 
                     nosim), 1, cfunc, 30)
        ),
  size = factor(rep(c(10, 20, n), rep(nosim, 3))))
g <- ggplot(dat, aes(x = x, fill = size)) + geom_histogram(binwidth=.3, colour = "black", aes(y = ..density..)) 
g <- g + stat_function(fun = dnorm, size = 2)
g + facet_grid(. ~ size)
```